---
title: "Analyzing a Simulated Labor Market"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing a Simulated Labor Market}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Goal

This vignette walks through a complete analysis of a simulated labor market:
simulate the data, explore it, estimate an AKM model, decompose wage variance,
and correct for limited mobility bias.

## 1. Simulate the Data

We simulate a labor market with 5,000 workers observed over 5 periods, with
moderate mobility (`lambda = 0.15`) and a Mincer wage equation:

```{r simulate}
library(labormarket)
library(data.table)

set.seed(2024)
lm_obj <- simlabormarket(
  nk = 4,       # 4 firm types
  nl = 5,       # 5 worker types
  nt = 5,       # 5 time periods
  ni = 5000,    # 5,000 workers
  lambda = 0.15, # 15% moving probability
  ratiog = 0.45, # 45% female
  mincer = TRUE   # Mincer wage equation
)

panel <- lm_obj@panel
```

## 2. Explore the Panel

### Dimensions

```{r dimensions}
cat("Observations:", nrow(panel), "\n")
cat("Workers:", length(unique(panel$i)), "\n")
cat("Firms:", length(unique(panel$fid)), "\n")
cat("Periods:", length(unique(panel$t)), "\n")
```

### Summary Statistics

```{r summary-stats}
# Wage and key variables
summary_vars <- panel[, .(
  mean_wage  = mean(lw),
  sd_wage    = sd(lw),
  mean_age   = mean(age),
  mean_educ  = mean(educ),
  mean_exper = mean(experience),
  pct_female = mean(gender)
)]
t(summary_vars)
```

### Wage Distribution

```{r wage-dist}
hist(panel$lw, breaks = 50, col = "#3B82A0", border = "white",
     main = "Distribution of Log Wages", xlab = "Log Wage")
```

### Gender Wage Gap

```{r gender-gap}
panel[, .(mean_wage = mean(lw), n = .N), by = gender]
```

### Movers vs Stayers

```{r movers}
# A mover is someone who changes fid across periods
movers <- panel[, .(n_firms = length(unique(fid))), by = i]
movers[, mover := n_firms > 1]
cat("Movers:", sum(movers$mover), "/", nrow(movers),
    "(", round(100 * mean(movers$mover), 1), "%)\n")
```

## 3. Estimate the AKM Model

We estimate the classic Abowd, Kramarz, and Margolis (1999) two-way fixed
effects decomposition:

$$\log w_{it} = \alpha_i + \psi_{j(i,t)} + \varepsilon_{it}$$

```{r akm}
library(fixest)

panel$fid_f <- as.factor(panel$fid)
panel$i_f   <- as.factor(panel$i)

# Restrict to the largest connected set
cf <- lfe::compfactor(list(panel$i_f, panel$fid_f))
largest <- names(which.max(table(cf)))
panel_conn <- panel[cf == largest]
panel_conn[, i_f   := droplevels(i_f)]
panel_conn[, fid_f := droplevels(fid_f)]

cat("Observations in connected set:", nrow(panel_conn), "/", nrow(panel), "\n")

# Estimate
model <- feols(lw ~ 1 | i_f + fid_f, data = panel_conn)
summary(model)
```

### Extract Fixed Effects

```{r fixed-effects}
fe <- fixef(model)
alpha_hat <- fe$i_f     # worker effects
psi_hat   <- fe$fid_f   # firm effects

cat("Worker effects — mean:", round(mean(alpha_hat), 3),
    " sd:", round(sd(alpha_hat), 3), "\n")
cat("Firm effects   — mean:", round(mean(psi_hat), 3),
    " sd:", round(sd(psi_hat), 3), "\n")
```

## 4. Variance Decomposition

The variance of log wages decomposes as:

$$\text{Var}(\log w) = \text{Var}(\hat\alpha) + \text{Var}(\hat\psi) + 2\,\text{Cov}(\hat\alpha, \hat\psi) + \text{Var}(\hat\varepsilon)$$

```{r variance-decomp}
# Match FEs back to observations
fe_list <- fixef(model)
# Use fixef_id to map observations to fixed effects (handles singletons)
alpha_vec <- fe_list$i_f[model$fixef_id$i_f]
psi_vec   <- fe_list$fid_f[model$fixef_id$fid_f]

# Subset to model observations if singletons were removed
obs_rm <- model$obs_selection$obsRemoved
if (!is.null(obs_rm)) {
  panel_conn <- panel_conn[obs_rm]  # negative indices = rows to drop
}

panel_conn[, alpha_hat := alpha_vec]
panel_conn[, psi_hat   := psi_vec]
panel_conn[, resid     := lw - alpha_hat - psi_hat]

var_wage  <- var(panel_conn$lw)
var_alpha <- var(panel_conn$alpha_hat)
var_psi   <- var(panel_conn$psi_hat)
cov_ap    <- cov(panel_conn$alpha_hat, panel_conn$psi_hat)
var_resid <- var(panel_conn$resid)

decomp <- data.frame(
  Component = c("Var(alpha)", "Var(psi)", "2*Cov(alpha,psi)", "Var(residual)", "Total"),
  Value = round(c(var_alpha, var_psi, 2 * cov_ap, var_resid, var_wage), 4),
  Share = round(c(var_alpha, var_psi, 2 * cov_ap, var_resid, var_wage) / var_wage * 100, 1)
)
decomp
```

### Worker–Firm Sorting

```{r sorting}
cor_ap <- cor(panel_conn$alpha_hat, panel_conn$psi_hat)
cat("Correlation(alpha, psi):", round(cor_ap, 3), "\n")
```

A positive correlation indicates assortative matching — high-type workers sort
into high-paying firms.

## 5. Correct Limited Mobility Bias

The variance decomposition above is **biased** when firms have few movers.
We use `lmbias()` to correct this:

```{r lmbias}
panel_conn$xbeta <- 0  # intercept-only model

results <- lmbias(model, panel_conn, R = 50)
print(round(results, 4))
```

### Interpretation

```{r interpret}
cat("--- Variance of Worker Effects ---\n")
cat("  Original: ", round(results["Original", 1], 4), "\n")
cat("  Bias:     ", round(results["Bias", 1], 4), "\n")
cat("  Corrected:", round(results["Corrected", 1], 4), "\n\n")

cat("--- Variance of Firm Effects ---\n")
cat("  Original: ", round(results["Original", 2], 4), "\n")
cat("  Bias:     ", round(results["Bias", 2], 4), "\n")
cat("  Corrected:", round(results["Corrected", 2], 4), "\n\n")

cat("--- Correlation ---\n")
cat("  Original: ", round(results["Original", 4], 4), "\n")
cat("  Corrected:", round(results["Corrected", 4], 4), "\n")
```

The bias correction typically **reduces** the estimated variance of both worker
and firm effects (since limited mobility inflates them) and can meaningfully
change the estimated sorting correlation.

## Summary

| Step | Function | What it does |
|------|----------|-------------|
| Simulate | `simlabormarket()` | Generate worker-firm panel with known DGP |
| Estimate | `fixest::feols()` | Two-way FE (AKM) decomposition |
| Correct | `lmbias()` | Bootstrap bias correction for limited mobility |

This workflow lets you study how parameters like mobility (`lambda`), number of
firm types (`nk`), and sample size (`ni`) affect the magnitude of limited
mobility bias — a key concern in empirical labor economics.
